Large Language Models in Reinforcement Learning - A Comparison of Methods

### 2023-11-09 doc/paper.tex
fix paper

Reinforcement Learning (RL) algorithms suffer when rewards are sparse and the state-action space is large. Even tasks which seem relatively simple can prove intractable if the completion of the task requires subtasks to be completed first, or if the reward only comes when the entire task has been completed. In such cases, random exploration is unlikely to lead the agent to discover a solution to the problem. The apparent simplicity of such problems is often due to human intuition, which allow us to quickly see possible solutions to a diverse set of problems. Large Language Models (LLMs) are trained on large corpora of human written text, and have been shown to capture parts of this intuition in many tasks. In recent years, LLMs have been successfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to. We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other.

@@

However, the causality behind which action is taken is different in LLM policy influencing; the LLM suggestions are added directly to the logits used to calculate probabilities of each action, meaning that the LLM directly influences the decision making process. Every time the LLM suggests performing a suboptimal action, this impacts the sampled action, regardless whether the PPO baseline has learned that another action is better or not. Over time, this may lead to the PPO not being able to explore other actions than the ones proposed by the LLM, thus making the PPO learn slower than if it was allowed to choose actions on its own. In other words, the LLM may have too much impact in the decision making when performing policy influencing, making the algorithm prone to errors from the LLM.

@@

>>>>

Reinforcement Learning (RL) algorithms suffer when rewards are sparse and the state-action space is large. Even tasks which seem relatively simple can prove intractable if the completion of the task requires subtasks to be completed first, or if the reward only comes when the entire task has been completed. In such cases, random exploration is unlikely to lead the agent to discover a solution to the problem. The apparent simplicity of such problems is often due to human intuition, which allow us to quickly see possible solutions to a diverse set of problems. Large Language Models (LLMs) are trained on large corpora of human written text, and have been shown to capture parts of this intuition in many tasks. In recent years, LLMs have been successfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to. We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other. \footnote{Code available at \href{https://github.com/jonatan-hanssen/in5490-project}{github}}

@@

However, the causality behind which action is taken is different in LLM policy influencing; the LLM suggestions are added directly to the logits used to calculate probabilities of each action, meaning that the LLM directly influences the decision making process. Every time the LLM suggests performing a suboptimal action, this impacts the sampled action, regardless whether the PPO baseline has learned that another action is better or not. Furthermore, as mentioned in Section \ref{emptyresultstext}, the logits increase in magnitude over time, while the contribution from the LLM does not scale with it. Since we cannot control the change in magnitude, this leads to unpredictable behavior from the model.

@@

A natural suggestion for how to improve the performance of both LLM policy influencing and A2C2 thus becomes to find a way to dynamically scale the influence from the LLM over time, so that the default scaling of logit contribution between the PPO and the LLM remains constant. This would allow for more predictability in which of the two contributes the most as learning progresses, and would make it possible to accurately customize how the rate of contribution from the LLM changes, for instance through simulated annealing.
